\section{Proving the semi\naive\ transformation correct}

\todo{rewrite from here on}

\label{section-logical-relation}

We have given two program transformations: $\phi e$, which optimizes $e$ by
computing fixed points semi\naive{}ly; and $\delta e$, which finds the change in
$\phi e$ under a change in its free variables.
%
To state the correctness of $\phi$ and $\delta$, we need to show that $\phi e$
preserves the meaning of $e$ and that $\delta e$ correctly updates $\phi e$ with
respect to changes in its variable bindings.
%
Since our transformations modify the types of higher-order expressions to
include the extra information needed for semi\naive\ evaluation, we cannot
directly prove that the semantics is preserved.
%
Instead, we formalize the relationship between $e$, $\phi e$, and $\delta e$
using a logical relation, and use this relation to prove an \emph{adequacy
  theorem} saying that the semantics is preserved for closed, first-order
programs.

\input{3-figure-logical-relation}

So, inductively on types $A$, letting $a,b \in \den{A}$, $x,y \in \den{\Phi A}$,
and $\dx \in \den{\DP A}$, we define a five place relation
$\weirdat{A}{\dx}{x}{a}{y}{b}$, meaning roughly ``$x,y$ speed up $a,b$
respectively, and $\dx$ changes $x$ into $y$''. The full definition is in
\cref{figure-seminaive-logical-relation}.

At product, sum, and function types this is essentially a more elaborate version
of the change structures given in \cref{section-change-structures}.
%
At set types, changes are still a set of values added to the initial value, but
we additionally insist that the ``slow'' $a,b$ and ``speedy'' $x,y$ are equal.
%
This is because we have engineered the definitions of $\Phi$ and $\phi$ to
preserve behavior on equality types.
%
Finally, since $\iso A$ represents values which cannot change, $\dx$ is an
uninformative empty tuple and the original and updated values are identical.
%
However, the ``speedy'' values are now \emph{pairs} of a value and its zero
change.
%
This ensures that at a boxed function type, we will always have a derivative (a
zero change) available.

The logical relation is defined on simple values, and so before we can state the
fundamental theorem, we have to extend it to contexts $\G$ and substitutions,
letting $\rho,\rho' \in \den{\G}$, $\g,\g' \in \den{\Phi\G}$, and $\dgamma \in
\den{\DP\G}$:

\nopagebreak[1]
\begin{align*}
  \weirdat{\G}{\dgamma}{\g}{\rho}{\g'}{\rho'}
  &\iff \fa{\hm x A \in \G}
  \weirdat{A}{\dgamma_{\mvar\dx}}{\g_{\mvar x}}{\rho_{\mvar x}}{\g'_{\mvar x}}{\rho'_{\mvar x}}
  \\
  &\hphantom{{}\iff{}}
  %\hspace*{-15.9pt} \wedge
  \hspace*{-18.6pt} \,\binwedge\,
  %\hspace*{-14.7pt} \binwedge
  \fa{\hd x A \in \G}
  \weirdat{\iso A}
          {\etuple{}}
          {(\g_{\dvar x}, \g_{\dvar \dx})}
          {\rho_{\dvar x}}
          {(\g'_{\dvar x}, \g'_{\dvar \dx})}
          {\rho'_{\dvar x}}
\end{align*}

\noindent
With that in place, we can state the fundamental theorem, showing that
$\phi$ and $\delta$ generate expressions which satisfy this logical
relation:

\begin{restatable}[Fundamental property]{theorem}{SeminaiveFundamental}
  \label{theorem-seminaive-fundamental}
  If $\J e \G A$ and $\weirdat{\G}{\dgamma}{\g}{\rho}{\g'}{\rho'}$ then
  \[\weirdat{A}{\den{\delta e} \<\tuple{\g,\dgamma}}{\den{\phi
      e}\<\g}{\den{e}\<\rho}{\den{\phi e}\<\g'}{\den{e}\<\rho'}\]
\end{restatable}

\nopagebreak[2]
\begin{proof}
  See \cref{proof-seminaive-fundamental}.
\end{proof}

\noindent
This theorem follows by a structural induction on typing derivations as usual,
but requires a number of lemmas.
%
By and large, these lemmas generalize or build on results stated earlier
regarding the simpler change structures from \cref{section-change-structures}.
%
For example, we build on \cref{lemma-phi-eqt,lemma-dummy-change} to
characterize the logical relation at equality types $\eqt A$ and the behavior of
\dummy:

%% \begin{lemma}[Equality Changes]
%%   If $\weirdat{\eqt A}{\dx}{x}{a}{y}{b}$ then $x = a$ and $y = b$.
%% \end{lemma}

%% \begin{lemma}[Dummy is zero at eqtypes]
%%   If $x \in \den{\eqt A}$ then $\weirdat{\eqt A}{\dummy\<x}{x}{x}{x}{x}$.
%% \end{lemma}

\begin{restatable}[Equality changes]{lemma}{EqualityChanges}
  \label{lemma-equality-changes}
  If $\weirdat{\eqt A}{\dx}{x}{a}{y}{b}$ then $x = a$ and $y = b$.
\end{restatable}
% TODO: why is restatable adding so much vertical space here?
\begin{restatable}[Dummy is zero at eqtypes]{lemma}{EqualityDummy}
  \label{lemma-equality-dummy}
  If $x \in \den{\eqt A}$ then $\weirdat{\eqt A}{\dummy\<x}{x}{x}{x}{x}$.
\end{restatable}

\begin{proof}
  In each case, induct on $\eqt A$. See
  \cref{proof-equality-changes,proof-equality-dummy}.
\end{proof}

\noindent
\Cref{lemma-equality-changes} tells us that at equality types, the sped-up version
of a value is the value itself. This is used later to prove our adequacy
theorem.
%
\Cref{lemma-equality-dummy} is an analogue of \cref{lemma-dummy-change}, showing
that \dummy\ function computes zero changes at equality types.
%
This is used in the proof of the fundamental theorem for \kw{for}-loops,
in whose $\phi$ and $\delta$ translations $\zero$ is implemented by \dummy.

Next, we generalize \cref{lemma-delta-lattice} to characterize changes at semilattice
type:

%% The lemma involves the lattice types, showing that a change for a lattice type
%% $L$ is something that can be joined on to it:

\begin{lemma}[Semilattice changes]
  At any semilattice type $L$, we have $\Delta L = L$, and moreover
  $\weirdat{L}{\dx}{x}{a}{y}{b}$ iff $x = a$ and $y = b = x \vee_L \dx$
\end{lemma}

\begin{proof}
  By induction on semilattice types $L$, applying \cref{lemma-equality-changes}
  (noting that every semilattice type is an equality type).
\end{proof}

\noindent
We require this lemma in the proofs of the fundamental theorem in all the
cases involving semilattice types -- namely $\bot$, ${\vee}$, \kw{for},
and \kw{fix}.

Since typing rules that involve discreteness (such as the $\iso$ rules)
manipulate the context, we need some lemmas regarding these manipulations.
First, we show that all valid changes for a context with only discrete variables
send substitutions to themselves, recalling that $\stripcx{\G}$ contains only the
discrete variables from $\G$.

\begin{restatable}[Discrete contexts don't change]{lemma}{DiscreteContexts}
  If $\weirdat{\stripcx{\G}}{\tuple{}}{\gamma}{\rho}{\gamma'}{\rho'}$ then
  $\gamma = \gamma'$ and $\rho = \rho'$.
\end{restatable}

\begin{restatable}{proof}{DiscreteContextsProof}
  All variables in the stripped contexts are discrete, and therefore the logical
  relation for discrete variables in contexts, which invokes the logical
  relation at $\iso$ type, requires their corresponding components be equal.
\end{restatable}

\noindent
We use this lemma in combination with the next, which says that any valid
context change gives rise to a valid change on a stripped context:

\begin{restatable}[Context stripping]{lemma}{ContextStripping}
  If $\weirdat{\G}{\dgamma}{\gamma}{\rho}{\gamma'}{\rho'}$
  then

  \[
  \weirdat
      {\stripcx{\G}}
      {\tuple{}}
      {\strip_{\Phi\G}(\g)}
      {\strip_{\G}(\rho)}
      {\strip_{\Phi\G}(\g')}
      {\strip_{\G}(\rho')}
  \]

  where $\strip_\G = \fork{\pi_{\dvar x}}_{\hd x A \in \G}$ keeps only the
  discrete variables from a substitution.
\end{restatable}

\begin{restatable}{proof}{ContextStrippingProof}
  Immediate from the definitions. \todo{shouldn't I be a little more explicit?}
\end{restatable}

\noindent
Jointly, these two lemmas ensure that a valid change to any context is an
identity on the discrete part. We use these in all the cases of the fundamental
theorem involving discrete expressions -- equality $\eeq{e_1}{e_2}$, set
literals $\esetsub{e_i}{i}$, emptiness tests $\eisempty e$, and box
introduction $\ebox e$.

Combining all these lemmas to establish the fundamental theorem, adequacy
follows immediately:

\begin{theorem}[Adequacy]
  If $\J e {\emptycx} {\eqt A}$ then $\den{e} = \den{\phi e}$.
\end{theorem}

\begin{proof}
  By \cref{theorem-seminaive-fundamental}, noting the premise is trivial since
  the context is empty, we have %
  $\weirdat{\eqt A}{\den{\delta e}}{\den{\phi e}}{\den{e}}{\den{\phi e}}{\den{e}}$%
  , which by \cref{lemma-equality-changes} implies $\den{\phi e} = \den{e}$.
\end{proof}
